---
title: "DAR3"
author: "Joshua Moffat"
date: "2025-04-16"
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, setup}
#After initial python environment is set-up, load reticulate library, assign python path, load remaining libraries
#Python path must be assigned right after loading reticulate, otherwise a temporary python environment will be created automatically

#Step 1
library(reticulate)

#Step 3
library(tidyverse); library(here)

# ----  Python environment only run once---------------------------------------------
# env <- "caption-env"
# if (!env %in% conda_list()$name) {
#   conda_create(env, packages = c("python=3.10"))
#   conda_install(env, c(
#     "tensorflow==2.11.0","numpy","pillow","tqdm",
#     "scikit-learn","matplotlib","nltk","pycocoevalcap","spacy"))
# }
#use_condaenv(env, required = TRUE)

#Step 2
python_path <- "/Users/joshuamoffat/Library/r-miniconda/envs/caption-env/bin/python"
use_python(python_path, required = TRUE)

#Run only once to make sure pandas and the correct version of tensorflow is installed
# try({
#   py_install(c("pandas", "tensorflow==2.11.0"), pip = TRUE)
# }, silent = TRUE)
```

```{python, check}
#Check important python packages have been installed
import tensorflow as tf
from tensorflow import keras
from tensorflow.keras import layers

print("✅ TensorFlow and Keras are ready!")
```


```{python, helpers}
#Import all python packages
import os, re, itertools, collections, json, math, random, pickle, warnings
from pathlib import Path
import numpy as np, pandas as pd
from PIL import Image
import tensorflow as tf, matplotlib.pyplot as plt
from tensorflow import keras; from tensorflow.keras import layers
from tqdm import tqdm
warnings.filterwarnings("ignore")

# ---------------------------------------------------------------------- #
#              Caption tokeniser (≥5 occurrences kept)                   #
# ---------------------------------------------------------------------- #
class CaptionTokenizer:
    def __init__(self, threshold=5):
        self.threshold = threshold
        self.word2idx = {"<pad>":0,"<start>":1,"<end>":2,"<unk>":3}
        self.idx2word = {v:k for k,v in self.word2idx.items()}

    def fit(self, captions):
        cnt = collections.Counter(itertools.chain(*captions))
        for w,f in cnt.items():
            if f>=self.threshold and w not in self.word2idx:
                i=len(self.word2idx)
                self.word2idx[w]=i; self.idx2word[i]=w

    def encode(self,toks):   return [self.word2idx.get(w,3) for w in toks]
    def decode(self,ids):    return [self.idx2word.get(i,"<unk>") for i in ids]
    def __len__(self):       return len(self.word2idx)
```

```{r, paths}
#Assign variables to the various file paths that we will access locally 
root <- here("flickr8k")        # <-- folder that holds images/ + captions.txt
img_dir <- file.path(root, "images")
capt_file <- file.path(root, "captions.txt")
stopifnot(file.exists(capt_file))

seed <- 42; set.seed(seed); reticulate::py_set_seed(seed)
output_dir <- here()                 # same folder as this Rmd file
```

```{python, clean}
#-------------------------------------------------------------
def clean(txt):
    txt = re.sub(r"[^\w\s]", "", txt.lower()).strip()
    return ["<start>"] + txt.split() + ["<end>"]

caps_raw = pd.read_csv(r.capt_file, sep=",", header=0)
caps_raw["tokens"] = caps_raw.caption.apply(clean)

# --------------  train / val / test split by image id --------------
files = caps_raw["image"].unique()
rng = np.random.default_rng(int(r.seed))
rng.shuffle(files)
n=len(files); train_ids, val_ids = files[:int(.7*n)], files[int(.7*n):int(.85*n)]
split = np.where(caps_raw.image.isin(train_ids),"train",
        np.where(caps_raw.image.isin(val_ids),"val","test"))
caps_raw["split"]=split

tok = CaptionTokenizer(threshold=5)
tok.fit(caps_raw.query("split=='train'")["tokens"])
vocab_size=len(tok)

PAD_TOKEN_ID = tok.word2idx["<pad>"]
END_TOKEN_ID = tok.word2idx["<end>"]
id2word      = tok.idx2word

# 95‑th percentile caption length
cap_len = caps_raw.tokens.map(len)
max_len = int(np.percentile(cap_len,95))
def encode_pad(toks, max_len=max_len):
    ids = tok.encode(toks)[:max_len]
    return ids + [0]*(max_len-len(ids))

caps_raw["enc"] = caps_raw.tokens.apply(encode_pad)

```

```{python, image}
#Smaller images size for faster runtime; downside is less information to train the CNN

IMG_SZ=224  #224 or 160

def load_img(path):
    img = Image.open(path).convert("RGB").resize((IMG_SZ,IMG_SZ))
    return np.asarray(img,dtype="float32")/255.0
  
 #Switch to batch 34 for to use less resources  
def make_ds(df, batch=64, shuffle=True):
    # Turn the dataframe into a tf.data.Dataset
    file_paths = [os.path.join(r.img_dir, fname) for fname in df["image"]]
    sequences  = list(df["enc"])
    labels     = [enc[1:] for enc in sequences]
    inputs     = [enc[:-1] for enc in sequences]

    ds = tf.data.Dataset.from_tensor_slices((file_paths, inputs, labels))

    # Shuffle before mapping for performance
    if shuffle:
        ds = ds.shuffle(buffer_size=len(df), seed=int(r.seed))

    def load_pair(path, seq_in, seq_out):
        img = tf.io.read_file(path)
        img = tf.image.decode_jpeg(img, channels=3)
        img = tf.image.resize(img, [IMG_SZ, IMG_SZ]) / 255.0
        return (img, seq_in), seq_out

    ds = ds.map(load_pair, num_parallel_calls=tf.data.AUTOTUNE)
    ds = ds.batch(batch).prefetch(tf.data.AUTOTUNE)
    return ds

  
  
```

```{python, imagedata}
#All 8k images used
train_ds = make_ds(caps_raw.query("split=='train'"))
val_ds   = make_ds(caps_raw.query("split=='val'"), shuffle=False)

#Subset data for debugging and more interactive/faster runs 
SUBSET = 3000 # number of caption rows to keep during grid search
VAL_SUBSET = 400 
train_subset = caps_raw.query("split=='train'").sample(SUBSET, random_state=int(r.seed))
val_subset   = caps_raw.query("split=='val'").sample(VAL_SUBSET, random_state=int(r.seed))   

train_ds = make_ds(train_subset, batch=64)      # faster batches
val_ds   = make_ds(val_subset,   batch=64)

#Check that the data set of batches has successfully populated
#Each batch is a combination images, caption_in and caption_out sequences of size X
for batch in train_ds.take(1):
    print("successfully loaded a training batch")

```


```{python, model}

class ReMask(layers.Layer):
    """Re‑attach the padding mask from the input seq_in onto the logits."""
    def call(self, inputs):
        logits, seq_in = inputs
        return logits

    def compute_mask(self, inputs, mask=None):
        # inputs = [logits, seq_in]; we ignore mask here
        seq_in = inputs[1]  # shape (batch, T)
        return tf.not_equal(seq_in, 0)  # True for real tokens, False for padding

def build_model(vocab_size, max_len,
                embed_dim   = 256,
                lstm_units  = 256,
                lr          = 1e-3,
                train_cnn   = False,
                unfreeze_at = -20):

    IMG_SZ = 224
    # ---------- CNN encoder ----------
    base = keras.applications.MobileNetV2(include_top=False,
                                          weights="imagenet",
                                          pooling="avg",
                                          input_shape=(IMG_SZ, IMG_SZ, 3))
    base.trainable = train_cnn
    if train_cnn and unfreeze_at < 0:          # optional fine‑tuning later
        for layer in base.layers[:unfreeze_at]:
            layer.trainable = False

    # ---------- inputs ----------
    IMG_SZ = 224 #224 or 160
    img_in = layers.Input((IMG_SZ, IMG_SZ, 3), name="image")
    seq_in = layers.Input((max_len - 1,), name="caption_in")  # minus 1 for image step

    # ---------- visual embedding ----------
    img_vec = layers.Dense(embed_dim, activation="relu")(base(img_in))   # (batch,E)

    # ---------- word embeddings WITH masking ----------
    tok_emb = layers.Embedding(vocab_size,
                               embed_dim,
                               mask_zero=True,
                               name='tok_emb')(seq_in)        # (batch, T, E)

    # ---------- add image token & merge masks ----------
    class AddImageToken(layers.Layer):
        def call(self, inputs):
            img, tok = inputs                 # img (batch,E)  tok (batch,T,E)
            img = tf.expand_dims(img, 1)      # (batch,1,E)
            return tf.concat([img, tok], axis=1)  # (batch,1+T,E)

        def compute_mask(self, inputs, mask=None):
            # mask[1] is token mask (batch,T); create (batch,1) True mask for image
            img_mask = tf.ones((tf.shape(inputs[0])[0], 1), dtype=tf.bool)
            return tf.concat([img_mask, mask[1]], axis=1)     # (batch,1+T)

    x = AddImageToken(name="add_img")([img_vec, tok_emb])                   # (batch,1+T,E)

    # ---------- LSTM decoder ----------
    x   = layers.LSTM(lstm_units, return_sequences=True)(x)
    
    # 1) drop the image‑token timestep
    x = layers.Cropping1D(cropping=(1, 0), name="slice_img")(x)
    
    # project to vocab
    logits = layers.TimeDistributed(layers.Dense(vocab_size), name="time_dist")(x)
    masked_logits = ReMask(name="re_mask")([logits, seq_in])
    
    # softmax + compile
    out = layers.Activation("softmax", name="softmax")(masked_logits)

    model = keras.Model([img_in, seq_in], out)
    model.compile(optimizer = keras.optimizers.Adam(lr),
                  loss      = "sparse_categorical_crossentropy")
    return model


```

```{python, modelattention}

import tensorflow as tf
from tensorflow import keras
from tensorflow.keras import layers

class ReMask(layers.Layer):
    """Re-attach the padding mask from seq_in onto the logits."""
    def call(self, inputs):
        logits, seq_in = inputs
        return logits
    def compute_mask(self, inputs, mask=None):
        seq_in = inputs[1]                       # (batch, T)
        return tf.not_equal(seq_in, 0)           # True for real tokens

def build_model(vocab_size, max_len,
                embed_dim   = 256,
                lstm_units  = 256,
                lr          = 1e-3,
                train_cnn   = False,
                unfreeze_at = -20):

    IMG_SZ = 224
    # -------- CNN encoder (no global pooling!) ---------------------
    base = keras.applications.MobileNetV2(include_top=False,
                                          weights="imagenet",
                                          pooling=None,             # keep spatial map
                                          input_shape=(IMG_SZ, IMG_SZ, 3))
    base.trainable = train_cnn
    if train_cnn and unfreeze_at < 0:
        for layer in base.layers[:unfreeze_at]:
            layer.trainable = False

    # -------- inputs ----------------------------------------------
    img_in = layers.Input((IMG_SZ, IMG_SZ, 3),  name="image")
    seq_in = layers.Input((max_len - 1,),        name="caption_in")  # T = max_len-1

    # -------- visual feature sequence  (S = 7*7 = 49) -------------
    feat_map  = base(img_in)                                         # (B,7,7,1280)
    feat_proj = layers.Conv2D(embed_dim, 1, activation="relu")(feat_map)
    feat_seq  = layers.Reshape((-1, embed_dim))(feat_proj)   # (B,S,E)
    
    
    # NEW: match depth to lstm_units so Attention sees (B,*,512)
    feat_seq  = layers.Dense(lstm_units, use_bias=False,
                         name="feat_to_hidden")(feat_seq)   # (B,49,U)

    # also build 1 global vector to init the LSTM
    img_vec   = layers.GlobalAveragePooling2D()(feat_map)            # (B,1280)
    init_h    = layers.Dense(lstm_units, activation="tanh")(img_vec) # (B,U)
    init_c    = layers.Dense(lstm_units, activation="tanh")(img_vec)

    # -------- word embeddings (masking on) ------------------------
    tok_emb = layers.Embedding(vocab_size,
                               embed_dim,
                               mask_zero=True,
                               name="tok_emb")(seq_in)               # (B,T,E)

    # -------- LSTM decoder ---------------------------------------
    dec_out = layers.LSTM(lstm_units,
                          return_sequences=True,
                          name="decoder")(tok_emb,
                                          initial_state=[init_h, init_c])    # (B,T,U)

    # -------- Luong / dot-product attention ----------------------
    #   Q = decoder timesteps  (B,T,U)
    #   K,V = visual regions   (B,S,E)      Attention does Q·K^T / sqrt(d_k)
    #ctx = layers.Attention(name="attn")([dec_out, feat_seq])         # (B,T,U)
    
    attn_layer = layers.Attention(name="attn")          # constructor clean
    ctx, attn_scores = attn_layer(                      # ask during call
        [dec_out, feat_seq],
        return_attention_scores=True)               # tuple  (B,T,U), (B,T,S)

    # concat context + decoder state, then project to vocab
    merged  = layers.Concatenate()([dec_out, ctx])                   # (B,T,2U)
    logits  = layers.TimeDistributed(layers.Dense(vocab_size),
                                     name="td_dense")(merged)
    masked  = ReMask(name="re_mask")([logits, seq_in])
    out     = layers.Activation("softmax", name="softmax")(masked)

    model = keras.Model([img_in, seq_in], out)
    model.compile(optimizer = keras.optimizers.Adam(lr),
                  loss      = "sparse_categorical_crossentropy")
        
    return model


```

```{python, callback SPICE}
        
from pycocoevalcap.spice.spice import Spice

class SpiceCallback(keras.callbacks.Callback):
    def __init__(self, ds, every=1, k=3):
        super().__init__()
        self.ds = ds
        self.every = every
        self.best = -1
        self.k = k

    def on_epoch_end(self, epoch, logs=None):
        if epoch % self.every != 0:
            return

        refs, hyps = {}, {}
        idx = 0

        for (imgs, seqs), labels in self.ds:
            expected_len = self.model.input[1].shape[1]  # e.g., 19
            seqs_fixed = tf.pad(seqs[:, :1], [[0, 0], [0, expected_len - 1]])
            preds = self.model.predict((imgs, seqs_fixed), verbose=0)
            pred_ids = np.argmax(preds[:, 0, :], axis=-1)

            for p, lab in zip(pred_ids, labels):
                # Decode label into list of tokens, remove <start>/<end>, then join
                decoded_ref_tokens = tok.decode(lab.numpy())[1:-1]
                ref_str = " ".join(decoded_ref_tokens)

                # Decode predicted token ID into word string
                pred_token = tok.idx2word.get(int(p), "<unk>")  # ensure p is an int
                hyp_str = pred_token

                refs[idx] = [ref_str]
                hyps[idx] = [hyp_str]
                idx += 1

        spice, _ = Spice().compute_score(refs, hyps)
        logs["val_spice"] = spice

        if spice > self.best:
            self.best = spice
            self.model.save(os.path.join(r.output_dir, "best_captioner.h5"))

        print(f"\nEpoch {epoch}: val_SPICE={spice:.3f}  (best={self.best:.3f})")

```

```{python, callbackSPICE2}
from pycocoevalcap.spice.spice import Spice
import numpy as np, tensorflow as tf, os, tqdm, keras

class SpiceCallback(keras.callbacks.Callback):
    def __init__(self, ds, tok, max_len, every=1, k=1, out_dir="ckpts"):
        super().__init__()
        self.ds      = ds
        self.tok     = tok
        self.max_len = max_len
        self.every   = every
        self.k       = k
        self.best    = -1
        self.out_dir = out_dir
        self.scorer  = Spice()          # keep JVM warm

    # ------------------------------------------------------------------
    # greedy decoder (k == 1)
    # ------------------------------------------------------------------
    def _greedy(self, img_tensor):
      seq = [self.tok.start_id]
      for _ in range(self.max_len - 1):
          pad_len  = self.max_len - len(seq)
          seq_in   = np.pad(seq, (0, pad_len))[: self.max_len - 1].astype("int32")

          img_b = img_tensor[None, ...]            # (1,H,W,3)
          seq_b = seq_in[None, ...]                # (1,T)
          logits = self.model.predict([img_b, seq_b], verbose=0)

          next_id = int(np.argmax(logits[0, len(seq)-1, :]))
          if next_id == self.tok.end_id:
              break
          seq.append(next_id)

    # 🔑  return plain string, not list
      return " ".join(self.tok.decode(seq[1:]))     # drop <start>


    # ------------------------------------------------------------------
    def on_epoch_end(self, epoch, logs=None):
        if epoch % self.every:
            return

        refs, hyps, idx = {}, {}, 0

        for (imgs, _), labels in tqdm.tqdm(self.ds, desc="SPICE   "):
            for img, lab in zip(imgs, labels):
                # -------- reference --------
                ref = " ".join(self.tok.decode(lab.numpy())[1:-1])
                refs[idx] = [ref]                     # list of refs

                # -------- hypothesis --------
                if self.k == 1:
                    hyp = self._greedy(img.numpy())
                else:
                    ids = beam_search_decode(self.model, img.numpy(),
                                             self.tok, self.max_len, k=self.k)
                    hyp = " ".join(self.tok.decode(ids[1:]))
                hyps[idx] = [hyp]                     # list **length 1**  ← FIX
                idx += 1

        spice, _ = self.scorer.compute_score(refs, hyps)
        logs["val_spice"] = spice

        if spice > self.best:
            self.best = spice
            self.model.save_weights(os.path.join(self.out_dir,
                                                 "best_captioner.h5"))

        print(f"\nEpoch {epoch}: val_SPICE = {spice:.4f}   (best = {self.best:.4f})")



```


```{python, beam4}
import os
import random
import json
import tempfile
import subprocess

import numpy as np
import tensorflow as tf
from tensorflow.keras.callbacks import Callback
from pycocoevalcap.spice.spice import Spice  # still imported, but unused
import pycocoevalcap.spice.spice as spice_module  # to find jar & cache

def beam_search_decode(model, img_tensor, tokenizer, max_len, k=3, alpha=0.7):
    beam = [([tokenizer.start_id], 0.0)]
    img_batch = img_tensor[None, ...]  # (1, H, W, 3)

    for _ in range(max_len - 1):
        new_beam = []
        for seq, log_p in beam:
            if seq[-1] == tokenizer.end_id:
                new_beam.append((seq, log_p))
                continue

            pad_len   = (max_len - 1) - len(seq)
            seq_pad   = seq + [tokenizer.pad_id] * pad_len
            seq_in    = tf.constant([seq_pad], dtype=tf.int32)    # (1, L)
            preds     = model([img_batch, seq_in], training=False)  # (1, L, V)

            idx_last  = len(seq) - 1
            logits    = preds[0, idx_last]
            log_probs = tf.math.log_softmax(logits)

            topk = tf.math.top_k(log_probs, k=k)
            for idx, lp_step in zip(topk.indices.numpy(), topk.values.numpy()):
                new_beam.append((seq + [int(idx)], log_p + float(lp_step)))

        new_beam.sort(key=lambda t: t[1], reverse=True)
        beam = new_beam[:k]
        if all(s[-1] == tokenizer.end_id for s, _ in beam):
            break

    best_seq = max(beam, key=lambda t: t[1] / (len(t[0]) ** alpha))[0]
    if best_seq and best_seq[0] == tokenizer.start_id:
        best_seq = best_seq[1:]
    if best_seq and best_seq[-1] == tokenizer.end_id:
        best_seq = best_seq[:-1]
    return tokenizer.decode(best_seq)


class SpiceCallback(Callback):
    def __init__(self, ds, tokenizer, max_len, output_dir,
                 every=1, k=3, eval_subset=None):
        super().__init__()
        self.ds           = ds
        self.tok          = tokenizer
        self.max_len      = max_len
        self.output_dir   = output_dir
        self.every        = every
        self.k            = k
        self.eval_subset  = eval_subset   # number of images to sample each epoch
        self.best         = -1.0

    def on_epoch_end(self, epoch, logs=None):
        if epoch % self.every != 0:
            print(f"[SPICE] skipping epoch {epoch}")
            return

        print(f"[SPICE] starting SPICE eval at epoch {epoch}")

        # 1) collect all (img, lab) pairs
        all_pairs = []
        for (img_batch, _), label_batch in self.ds:
            for img, lab in zip(img_batch, label_batch):
                all_pairs.append((img, lab))

        # 2) shuffle & take subset if requested
        if self.eval_subset is not None and len(all_pairs) > self.eval_subset:
            random.shuffle(all_pairs)
            pairs = all_pairs[:self.eval_subset]
        else:
            pairs = all_pairs

        # 3) decode into refs & hyps (plain strings)
        refs, hyps = {}, {}
        for idx, (img, lab) in enumerate(pairs):
            hyp = beam_search_decode(self.model, img, self.tok, self.max_len, k=self.k)
            ref_tokens = self.tok.decode(lab.numpy())[1:-1]
            refs[idx] = [" ".join(ref_tokens)]
            hyps[idx] = hyp

            if (idx + 1) % 10 == 0:
                print(f"[SPICE]  – processed {idx+1} images")

        print(f"[SPICE] total images processed: {len(pairs)}, computing SPICE…")

        # 4) build JSON payload with test as plain string
        payload = []
        for img_id, ref_list in refs.items():
            raw_hyp = hyps[img_id]
            test_str = raw_hyp[0] if isinstance(raw_hyp, (list, tuple)) else raw_hyp
            payload.append({
                "image_id": img_id,
                "refs":     ref_list,
                "test":     test_str
            })

        # 5) dump to temp file and call the JAR
        # inp = tempfile.NamedTemporaryFile(delete=False, suffix=".json")
        #inp = tempfile.NamedTemporaryFile(mode='w', delete=False, suffix=".json")
        #json.dump(payload, inp)
        #inp.close()
        #outp = inp.name + ".out"
       
        with tempfile.NamedTemporaryFile(mode='w', delete=False, suffix=".json") as inp:
            json.dump(payload, inp)
            inp_path = inp.name
        outp = inp_path + ".out"

        base_dir = os.path.dirname(spice_module.__file__)
        jar_path = os.path.join(base_dir, "spice-1.0.jar")
        cache    = os.path.join(base_dir, "cache")

        cmd = [
            "java", "-jar", jar_path,
            inp.name,
            "-cache", cache,
            "-out", outp,
            "-silent"
        ]
      
        subprocess.check_call(cmd)

        # 6) read back and extract the score
        with open(outp, "r") as f:
            result = json.load(f)
            
        # if the JAR returned a list, take its first element:
        if isinstance(result, list):
          result = result[0]
        
        #print("[SPICE] raw JAR output:", json.dumps(result, indent=2))  
        #spice_score = result["scores"]["All"]["score"]
        
        scores_dict = result.get("scores", {})
        all_section = scores_dict.get("All", {})

        # Pull out the SPICE F-score
        spice_score = all_section.get("f")
        if spice_score is None:
          raise KeyError(f"Couldn't find 'f' in SPICE output: {scores_dict}")

        # 7) log & checkpoint
        logs = logs or {}
        logs["val_spice"] = spice_score

        if spice_score > self.best:
            print(f"[SPICE] 🎉 new best ({spice_score:.4f}), saving model")
            self.best = spice_score
            os.makedirs(self.output_dir, exist_ok=True)
            self.model.save(os.path.join(self.output_dir, "best_captioner.h5"))
        else:
            print(f"[SPICE] best remains {self.best:.4f}")

        print(f"\nEpoch {epoch}: val_SPICE = {spice_score:.4f} (best = {self.best:.4f})")

```


```{python, computeSPICE}
from pycocoevalcap.spice.spice import Spice

def compute_spice(model, ds, tok, k=3):
    refs, hyps = {}, {}
    idx = 0

    for (imgs, seqs), labels in ds:
        expected_len = model.input[1].shape[1]
        seqs_fixed = tf.pad(seqs[:, :1], [[0, 0], [0, expected_len - 1]])
        preds = model.predict((imgs, seqs_fixed), verbose=0)
        pred_ids = np.argmax(preds[:, 0, :], axis=-1)

        for p, lab in zip(pred_ids, labels):
            decoded_ref_tokens = tok.decode(lab.numpy())[1:-1]
            ref_str = " ".join(decoded_ref_tokens)

            pred_token = tok.idx2word.get(int(p), "<unk>")
            hyp_str = pred_token

            refs[idx] = [ref_str]
            hyps[idx] = [hyp_str]
            idx += 1

    spice, _ = Spice().compute_score(refs, hyps)
    return spice

```

```{python, computeSPICE2}

from pycocoevalcap.spice.spice import Spice
import numpy as np, tensorflow as tf

# ------------------------------------------------------------------
# simple greedy decoder  (fixed) -----------------------------------
# ------------------------------------------------------------------
def greedy_decode(model, img, tok, max_len):
    """Return a plain-string caption generated with greedy decoding."""
    seq = [tok.start_id]
    for _ in range(max_len - 1):
        pad     = max_len - len(seq)
        seq_in  = np.pad(seq, (0, pad))[: max_len - 1].astype("int32")  # (T,)

        img_b = img[None, ...]            # (1,H,W,3)
        seq_b = seq_in[None, ...]         # (1,T)
        logits = model.predict([img_b, seq_b], verbose=0)

        next_id = int(np.argmax(logits[0, len(seq)-1, :]))
        if next_id == tok.end_id:
            break
        seq.append(next_id)

    # ---- return single caption string (not list) ------------------
    return " ".join(tok.decode(seq[1:]))     # drop <start>

# ------------------------------------------------------------------
def compute_spice(model, ds, tok, max_len, k=1):
    """
    Compute SPICE over dataset `ds`.
    If k==1 use greedy; if k>1 call beam_search_decode().
    """
    refs, hyps, idx = {}, {}, 0
    scorer = Spice()                         # reuse JVM

    for (imgs, _), labels in ds:
        for img, lab in zip(imgs, labels):
            print(f"Processing example {idx}...")
            # -------- reference --------
            ref = " ".join(tok.decode(lab.numpy())[1:-1])   # strip <start>/<end>
            refs[idx] = [ref]                               # list of refs

            # -------- hypothesis -------
            if k == 1:
                hyp = greedy_decode(model, img.numpy(), tok, max_len)
            else:
                ids = beam_search_decode(model, img.numpy(),
                                          tok, max_len, k=k)
                hyp = " ".join(tok.decode(ids[1:]))
            hyps[idx] = [hyp]                               # list length 1

            idx += 1

    spice, _ = scorer.compute_score(refs, hyps)
    return spice


```



```{python, computespicebeam2}

import os
import json
import tempfile
import subprocess

import numpy as np
import tensorflow as tf

# to locate the JAR and cache directory
import pycocoevalcap.spice.spice as spice_module

def generate_caption_beam(model, image, tokenizer, max_len=20, beam_width=3, alpha=0.7):
    beam = [([tokenizer.start_id], 0.0)]
    img_batch = image[None, ...]
    for _ in range(max_len - 1):
        candidates = []
        for seq, score in beam:
            if seq[-1] == tokenizer.end_id:
                candidates.append((seq, score))
                continue
            pad_len = (max_len - 1) - len(seq)
            seq_pad = seq + [tokenizer.pad_id] * pad_len
            seq_in  = tf.constant([seq_pad], dtype=tf.int32)
            preds   = model([img_batch, seq_in], training=False)
            logits  = preds[0, len(seq)-1]
            logp    = tf.nn.log_softmax(logits).numpy()
            topk    = np.argsort(logp)[-beam_width:]
            for idx in topk:
                candidates.append((seq + [int(idx)], score + float(logp[idx])))
        candidates.sort(key=lambda t: t[1], reverse=True)
        beam = candidates[:beam_width]
        if all(s[-1] == tokenizer.end_id for s,_ in beam):
            break

    best_seq = max(beam, key=lambda t: t[1]/(len(t[0])**alpha))[0]
    if best_seq and best_seq[0] == tokenizer.start_id:
        best_seq = best_seq[1:]
    if best_seq and best_seq[-1] == tokenizer.end_id:
        best_seq = best_seq[:-1]
    return tokenizer.decode(best_seq)

def compute_spice(model, ds, tokenizer, max_len=20, beam_width=3, eval_subset=None):
    # 1) collect (img, lab) pairs
    all_pairs = []
    for (imgs, _), labs in ds:
        for img, lab in zip(imgs, labs):
            all_pairs.append((img, lab))

    # 2) optionally sample a subset
    if eval_subset and len(all_pairs) > eval_subset:
        np.random.shuffle(all_pairs)
        all_pairs = all_pairs[:eval_subset]

    # 3) decode all captions
    refs, hyps = {}, {}
    for idx, (img, lab) in enumerate(all_pairs):
        hyp = generate_caption_beam(model, img, tokenizer,
                                   max_len=max_len, beam_width=beam_width)
        ref_tokens = tokenizer.decode(lab.numpy())[1:-1]
        refs[idx] = [" ".join(ref_tokens)]
        hyps[idx] = hyp

    # 4) build JSON payload with test as string
    payload = []
    for img_id, ref_list in refs.items():
        raw_hyp = hyps[img_id]
        test_str = raw_hyp[0] if isinstance(raw_hyp, (list, tuple)) else raw_hyp
        payload.append({
            "image_id": img_id,
            "refs":     ref_list,
            "test":     test_str
        })

    # 5) write to temp file & call SPICE JAR
    with tempfile.NamedTemporaryFile(mode='w', delete=False, suffix='.json') as inp:
        json.dump(payload, inp)
        inp_path = inp.name
    out_path = inp_path + '.out'

    base_dir = os.path.dirname(spice_module.__file__)
    jar_path = os.path.join(base_dir, 'spice-1.0.jar')
    cache    = os.path.join(base_dir, 'cache')
    cmd = [
        'java', '-jar', jar_path,
        inp_path,
        '-cache', cache,
        '-out',   out_path,
        '-silent'
    ]
    subprocess.check_call(cmd)

    # 6) read back & extract the F-score
    with open(out_path, 'r') as f:
        result = json.load(f)
    if isinstance(result, list) and result:
        result = result[0]
    all_scores = result.get('scores', {}).get('All', {})
    if 'f' not in all_scores:
        raise KeyError(f"No 'f' in SPICE output: {all_scores}")
    return all_scores['f']


```

```{python, mask_probe}
import sys

grid = [
    dict(embed_dim=128, lstm_units=256, lr=1e-3)
]

histories = []
MAX_EPOCHS = 2

for i, cfg in enumerate(grid, 1):
    print(f"\n▶▶  Hyper‑run {i}/{len(grid)}  —  {cfg}")
    try:
        # ----- build & compile model -----
        model = build_model(
            vocab_size, max_len,
            embed_dim = cfg["embed_dim"],
            lstm_units = cfg["lstm_units"],
            lr        = cfg["lr"],
            train_cnn = False          # keep CNN frozen for speed
        )
        print("Built model")
        
                # ----- MASK‑PROBE: inspect masks before training -----
        from tensorflow.keras import Model

        # 1) get inputs
        img_in  = model.get_layer("image").input
        seq_in  = model.get_layer("caption_in").input

        # 2) get the three key intermediate outputs
        tok_emb_out  = model.get_layer("tok_emb").output
        add_img_out  = model.get_layer("add_img").output
        slice_out    = model.get_layer("slice_img").output
        post_mask_out = model.get_layer("re_mask").output

        # 3) build a small probe model
        probe = Model([img_in, seq_in],
                      [tok_emb_out, add_img_out, slice_out, post_mask_out])

        # 4) grab a single batch correctly:
        (img_batch, seq_batch), _ = next(iter(train_ds))
        #    img_batch.shape → (batch,160,160,3)
        #    seq_batch.shape → (batch, max_len-1)

        # 5) compute and print the masks
        masks = probe.compute_mask([img_batch, seq_batch],mask=[None, None])
        print("Mask after Embedding:     ", masks[0].numpy())
        print("Mask after AddImageToken:  ", masks[1].numpy())
        print("Mask after Slicing:        ", masks[2])  # should be None if lost
        print("After post_mask:",       masks[3].numpy()) # now a boolean array
        # ----- end MASK‑PROBE -----


        # ----- callbacks -----
        callbacks = [
            SpiceCallback(val_ds, every=MAX_EPOCHS),
            keras.callbacks.EarlyStopping(
                monitor="val_loss",
                patience=2,
                mode="min",
                restore_best_weights=True
            )
        ]
        
        print("Start training")

        # ----- training -----
        hist = model.fit(
            train_ds,
            validation_data = val_ds,
            epochs          = MAX_EPOCHS, #starting with few epochs for computational complexity 
            callbacks       = callbacks,
            verbose         = 1
        )
        
        print("Done training")

        # ----- record results -----
        cfg["best_spice"] = max(hist.history.get("val_spice", [0]))
        cfg["hist"]       = hist.history
        histories.append(cfg)

    except Exception as e:
        print(f"❌ run failed: {e}")     # keeps the grid search moving
        continue

if not histories:                                 # all runs failed
    raise RuntimeError("All grid runs failed – check earlier errors.")

# grab one example from your batch
(seq, mask_row) = seq_batch[0], masks[3][0].numpy()

print("Sequence:", seq.numpy())
print("Mask:    ", mask_row)

#You should see zeros in Sequence exactly where Mask is False.

```

We batch variable‐length captions by zero‑padding shorter sequences—but since those zeros aren’t real tokens, we don’t want the model to predict or be penalized on them. By using `mask_zero=True` in the Embedding, merging and cropping that mask through our custom `AddImageToken` and `Cropping1D` layers, then re‑attaching the original `(seq_in != 0)` mask with a lightweight `ReMask` layer before softmax, we ensure `sparse_categorical_crossentropy` only sums over valid timesteps. This prevents padding tokens from corrupting the loss and focuses training on real caption data.


```{python, debugg}

def decode_batch(logits, id2word, end_token_id=None, pad_token_id=0):
    """
    logits:   np.array or tf.Tensor of shape (batch, T, vocab_size)
    id2word:  dict mapping token_id → word string
    end_token_id:  integer ID of your <end> token (if you have one)
    pad_token_id:  integer ID of your padding token (often 0)
    """
    # 1) pick the most likely token at each timestep
    token_ids = np.argmax(logits, axis=-1)  # shape (batch, T)
    
    captions = []
    for seq in token_ids:
        words = []
        for idx in seq:
            if end_token_id is not None and idx == end_token_id:
                break
            if idx == pad_token_id:
                break
            words.append(id2word.get(idx, "<unk>"))
        captions.append(" ".join(words))
    return captions

```


```{python, gridsearch}
# -------------------------------------------------------------------
#  Grid search loop  (4 configs)  with robust try/except protection
# -------------------------------------------------------------------
#from tensorflow import keras
import sys

#grid search w/ 8 configs
grid = [
    dict(embed_dim=e, lstm_units=l, lr=a)
    for e in (128,256) for l in (256,512) for a in (1e-3,1e-4)
]

grid = [
    dict(embed_dim=256, lstm_units=512, lr=1e-3)
   
]

grid = [
    dict(embed_dim=128, lstm_units=256, lr=1e-3),
    dict(embed_dim=256, lstm_units=256, lr=1e-3),
    dict(embed_dim=256, lstm_units=512, lr=1e-4),
    dict(embed_dim=128, lstm_units=512, lr=1e-4)
]

histories = []
MAX_EPOCHS = 5

for i, cfg in enumerate(grid, 1):
    print(f"\n▶▶  Hyper‑run {i}/{len(grid)}  —  {cfg}")
    try:
        # ----- build & compile model -----
        model = build_model(
            vocab_size, max_len,
            embed_dim = cfg["embed_dim"],
            lstm_units = cfg["lstm_units"],
            lr        = cfg["lr"],
            train_cnn = False          # keep CNN frozen for speed
        )
        print("Built model")

        # ----- callbacks ----- turn off spice call back for speed; call final spice at the end
        callbacks = [
          
            keras.callbacks.ReduceLROnPlateau(
                monitor="val_loss",   # same metric as EarlyStopping
                factor=0.5,           # cut LR in half
                patience=1,           # after 1 epoch with no val‑loss improvement
                min_lr=1e-6,          # don’t let it go to zero
                verbose=1
            ),
            #SpiceCallback(val_ds, every=1), 
            keras.callbacks.EarlyStopping(
                monitor="val_loss",
                patience=2,
                mode="min",
                restore_best_weights=True
            )
        ]
        
        print("Start training")

        # ----- training -----
        hist = model.fit(
            train_ds,
            validation_data = val_ds,
            epochs          = MAX_EPOCHS, #starting with few epochs for computational complexity 
            callbacks       = callbacks,
            verbose         = 1
        )
        
        # ----- DEBUGGING SPICE ZERO ----- 
        # grab one batch from your validation set
        (img_batch, seq_batch), _ = next(iter(val_ds))
        # run a forward pass
        logits = model.predict([img_batch, seq_batch])
        token_ids = np.argmax(logits, axis=-1)
        print("Unique predicted IDs:", np.unique(token_ids))
        print("ID counts:", dict(zip(*np.unique(token_ids, return_counts=True))))
        print("First sequence token IDs:", token_ids[0])
        
        # ensure it’s a numpy array
        if isinstance(token_ids, tf.Tensor):
          token_ids = token_ids.numpy()

        mapped = []
        for idx in token_ids[0]:
          i = int(idx)
          if i == PAD_ID or i == END_ID:
              break
        mapped.append(id2word.get(i, "<unk>"))
        print("Mapped words:", mapped)
        
        # decode logits → token IDs → words (you’ll already have a function for this)
        decoded = decode_batch(logits, id2word, end_token_id=END_TOKEN_ID, pad_token_id=PAD_TOKEN_ID)  
        print("🔍 Sample decoded captions (first 5):")
        for cap in decoded[:5]:
          print(" •", cap)

        print("Done training and debug output")
        # ----- end DEBUGGING -----
        
        print("Done training")

        # ----- record results -----
        cfg["best_spice"] = max(hist.history.get("val_spice", [0]))
        
        spice_score = compute_spice(model, val_ds, tok)
        cfg["final_spice"] = spice_score
        cfg["val_loss"] = min(hist.history["val_loss"])
        cfg["hist"] = hist.history
        histories.append(cfg)
        print(f"\n✅ Final SPICE score: {spice_score:.4f}")


    except Exception as e:
        print(f"❌ run failed: {e}")     # keeps the grid search moving
        continue

# -------------------------------------------------------------------
#  Pick the best configuration
# -------------------------------------------------------------------
if not histories:                                 # all runs failed
    raise RuntimeError("All grid runs failed – check earlier errors.")

# best_cfg = max(histories, key=lambda d: d["best_spice"])
# print("\n🏆  Best config:", best_cfg)
#best_cfg = max(histories, key=lambda d: d["final_spice"])
best_cfg = min(histories, key=lambda d: d["val_loss"])
print("\n🏆  Best config:", best_cfg)

# (Optional) save results
json.dump(histories,
          open(os.path.join(r.output_dir, "grid_results.json"), "w"),
          indent=2)

```

Select the top 2–3 configs, not just the single best one

1st embed_dim: 256 | lstm_units: 512 | lr: 0.001 - val_loss: 3.6345
2nd embed_dim: 128 | lstm_units: 512 | lr: 0.001 - val_loss: 3.7287
3rd embed_dim: 256 | lstm_units: 256 | lr: 0.001 - val_loss: 3.8715
During the full‑data run:

Save both best‑epoch and final checkpoints (ModelCheckpoint(save_best_only=True, monitor="val_spice") plus EarlyStopping).

Compare their final validation SPICE—sometimes the best‑epoch model edges out by 0.2–0.5 points.

Pick the checkpoint with the higher validation SPICE on the full dataset, then compute test‑set metrics.

```{python, winner}

full_configs = [
    dict(embed_dim=256, lstm_units=512, lr=1e-3),
    dict(embed_dim=128, lstm_units=512, lr=1e-3),
    dict(embed_dim=256, lstm_units=256, lr=1e-3),
]

# ------------------------------------------------------------------
# Ensure the checkpoint directory exists
# ------------------------------------------------------------------
Path("ckpts").mkdir(parents=True, exist_ok=True)   # does nothing if already there

results = []
for j, cfg in enumerate(full_configs, 1):

    tag = f"full{j}_ed{cfg['embed_dim']}_lu{cfg['lstm_units']}"
    model = build_model(vocab_size, max_len, **cfg, train_cnn=False)
    
    spice_cb  = SpiceCallback(val_ds, every=1)

    best_ckpt  = keras.callbacks.ModelCheckpoint(
        f"ckpts/{tag}_best.weights.h5",
        monitor="val_spice",
        mode="max",
        save_best_only=True,
        save_weights_only=True,
        verbose=1
    )
    early_stop = keras.callbacks.EarlyStopping(
        monitor="val_loss",
        patience=3,
        restore_best_weights=True
    )
    reduce_lr  = keras.callbacks.ReduceLROnPlateau(
        monitor="val_loss",
        factor=0.5,
        patience=1,
        min_lr=1e-6,
        verbose=1
    )

    hist = model.fit(
        train_ds,
        validation_data = val_ds,
        epochs          = 30,          # plenty; ES will stop sooner
        callbacks       = [spice_cb, best_ckpt, early_stop, reduce_lr],
        verbose         = 1
    )
    
    # ----- ② save EarlyStopping‑restored “final” weights -----------
    final_path = f"ckpts/{tag}_final.weights.h5"
    model.save_weights(final_path)
    print(f"Saved final weights → {final_path}")

    # -----------------------------------------------------------------
    # Evaluate both checkpoints on the validation set
    # -----------------------------------------------------------------
    model.load_weights(f"ckpts/{tag}_best.weights.h5")
    spice_best = compute_spice(model, val_ds, tok)

    model.load_weights(final_path)  # ES-restored
    spice_final = compute_spice(model, val_ds, tok)

    results.append(dict(cfg=cfg,
                        spice_best = spice_best,
                        spice_final= spice_final,
                        val_loss   = min(hist.history['val_loss']),
                        history = hist.history
                        ))
                        
                        
winner = max(results, key=lambda d: max(d['spice_best'], d['spice_final']))
print("🏆  Winning config:", winner['cfg'])
print(f"   SPICE  best‑epoch : {winner['spice_best']:.6f}")
print(f"   SPICE  final‑epoch: {winner['spice_final']:.6f}")

#fhistory = winner["history"]

json.dump(results,
          open(os.path.join(r.output_dir, "grid_results2.json"), "w"),
          indent=2)
          
Winning config: {'embed_dim': 128, 'lstm_units': 512, 'lr': 0.001}
```
Winning config: {'embed_dim': 128, 'lstm_units': 512, 'lr': 0.001} for SPICE
```{python}

# --- rebuild exactly the same architecture ---
fbest_cfg = winner["cfg"]          # already printed above
modelb = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=False)

# choose which checkpoint you want to showcase
ckpt_path = f"ckpts/full2_ed128_lu512_best.weights.h5"
# if you’d rather show the Early-Stopping “final” weights, swap the filename here
modelb.load_weights(ckpt_path)
print("✓  Weights loaded:", ckpt_path)


def preprocess_image(path, target_size=160):
    """Read & preprocess 1 image exactly as in training."""
    img_raw = tf.io.read_file(path)
    img     = tf.image.decode_jpeg(img_raw, channels=3)
    img     = tf.image.resize(img, (target_size, target_size)) / 255.0
    return img

import tensorflow as tf
from typing import List, Tuple

def beam_search_decode(model, img_tensor, tokenizer, max_len, k=3, alpha=0.7):
    beam = [([tokenizer.start_id], 0.0)]           # (tokens, log-prob)
    img_batch = img_tensor[None, ...]              # (1, H, W, 3)

    for _ in range(max_len - 1):
        new_beam = []

        for seq, log_p in beam:
            if seq[-1] == tokenizer.end_id:        # branch already finished
                new_beam.append((seq, log_p))
                continue

            # ---------- pad to fixed length ----------
            pad_len  = (max_len - 1) - len(seq)
            seq_pad  = seq + [tokenizer.pad_id] * pad_len
            seq_in   = tf.constant([seq_pad], dtype=tf.int32)   # (1, L)
            preds    = model([img_batch, seq_in])               # (1, L, V)

            # take logits at position len(seq)-1 (current last real token)
            idx_last = len(seq) - 1
            logits   = preds[0, idx_last]
            log_probs = tf.math.log_softmax(logits)

            # ---------- expand top-k ----------
            topk = tf.math.top_k(log_probs, k=k)
            for idx, lp_step in zip(topk.indices.numpy(), topk.values.numpy()):
                new_seq   = seq + [int(idx)]
                new_log_p = log_p + float(lp_step)
                new_beam.append((new_seq, new_log_p))

        new_beam.sort(key=lambda t: t[1], reverse=True)
        beam = new_beam[:k]

        if all(s[-1] == tokenizer.end_id for s, _ in beam):
            break

    # ---------- length-normalised best ----------
    best_seq = max(beam, key=lambda t: t[1] / (len(t[0]) ** alpha))[0]

    # strip special tokens
    if best_seq[0] == tokenizer.start_id: best_seq = best_seq[1:]
    if best_seq and best_seq[-1] == tokenizer.end_id: best_seq = best_seq[:-1]

    return tokenizer.decode(best_seq)



def show_caption(path, caption):
    img_disp = Image.open(path)
    plt.imshow(img_disp); plt.axis('off')
    plt.title(caption, fontsize=12, wrap=True)
    plt.show()
    
    
# ——— tokenizer patch (only needed once) ———
tok.start_id = tok.word2idx["<start>"]
tok.end_id   = tok.word2idx["<end>"]
tok.pad_id   = 0

# grab a handful of validation-set file paths
val_files = val_subset["image"].sample(5, random_state=42)
# (or use test images you’ve never seen)

for fname in val_files:
    path = os.path.join(r.img_dir, fname)
    img_tensor = preprocess_image(path)
    caption    = beam_search_decode(modelb, img_tensor, tok, max_len, k=3)
    show_caption(path, caption)

```

```{python, finalhyper}

final_config = [dict(embed_dim=128, lstm_units=512, lr=1e-3)]
# ------------------------------------------------------------------
# Ensure the checkpoint directory exists
# ------------------------------------------------------------------
Path("ckpts_f").mkdir(parents=True, exist_ok=True)   # does nothing if already there

fresults = []

for j, cfg in enumerate(final_config, 1):

    tag = f"full{j}_ed{cfg['embed_dim']}_lu{cfg['lstm_units']}"
    model = build_model(vocab_size, max_len, **cfg, train_cnn=False)
    
    spice_cb  = SpiceCallback(val_ds, every=1)
    #spice_cb = SpiceCallback(ds=val_ds, tokenizer=tok, max_len=20, output_dir=r.output_dir, every=1, k=3, eval_subset=50)  #k=beam width

    best_ckpt  = keras.callbacks.ModelCheckpoint(
        f"ckpts_f/{tag}_best.weights.h5",
        monitor="val_spice",
        mode="max",
        save_best_only=True,
        save_weights_only=True,
        verbose=1
    )
    early_stop = keras.callbacks.EarlyStopping(
        monitor="val_loss",
        patience=20,
        restore_best_weights=True
    )
    reduce_lr  = keras.callbacks.ReduceLROnPlateau(
        monitor="val_loss",
        factor=0.5,
        patience=5,
        min_lr=1e-6,
        verbose=1
    )

    hist = model.fit(
        train_ds,
        validation_data = val_ds,
        epochs          = 50,          # plenty; ES will stop sooner
        callbacks       = [spice_cb, best_ckpt, early_stop, reduce_lr],
        verbose         = 1
    )
    
    # ----- ② save EarlyStopping‑restored “final” weights -----------
    final_path = f"ckpts_f/{tag}_final.weights.h5"
    model.save_weights(final_path)
    print(f"Saved final weights → {final_path}")

    # -----------------------------------------------------------------
    # Evaluate both checkpoints on the validation set
    # -----------------------------------------------------------------
    model.load_weights(f"ckpts_f/{tag}_best.weights.h5")
    spice_best = compute_spice(model, val_ds, tok)
    #spice_best = compute_spice(model, val_ds, tok, max_len=20, beam_width=3, eval_subset=50)

    model.load_weights(final_path)  # ES-restored
    spice_final = compute_spice(model, val_ds, tok)
    #spice_final = compute_spice(model, val_ds, tok, max_len=20, beam_width=3, eval_subset=50)

    fresults.append(dict(cfg=cfg,
                        spice_best = spice_best,
                        spice_final= spice_final,
                        val_loss   = min(hist.history['val_loss']),
                        history = hist.history
                        ))
                        

```

```{python, finetune}
final_config = [dict(embed_dim=128, lstm_units=512, lr=1e-3)]

Path("ckpts_f3").mkdir(parents=True, exist_ok=True)

f3results = []

for j, cfg in enumerate(final_config, 1):

    tag = f"full{j}_ed{cfg['embed_dim']}_lu{cfg['lstm_units']}"
    model = build_model(vocab_size, max_len, **cfg, train_cnn=False)
    
    val_subs = val_ds.take(10)
    spice_cb  = SpiceCallback(
      ds        = val_subs,
      tok       = tok,
      max_len   = max_len,
      every     = 1,      # compute each epoch
      k         = 1,      # set >1 to switch to beam
      out_dir   = "ckpts_f3"
    )

    best_ckpt  = keras.callbacks.ModelCheckpoint(
        f"ckpts_f3/{tag}_best.weights.h5",
        monitor="val_spice",
        mode="max",
        save_best_only=True,
        save_weights_only=True,
        verbose=1
    )
    early_stop = keras.callbacks.EarlyStopping(
        monitor="val_loss",
        patience=15,
        restore_best_weights=True
    )
    reduce_lr  = keras.callbacks.ReduceLROnPlateau(
        monitor="val_loss",
        factor=0.5,
        patience=4,
        min_lr=1e-6,
        verbose=1
    )

    hist = model.fit(
        train_ds,
        validation_data = val_ds,
        epochs          = 50,          
        callbacks       = [spice_cb, best_ckpt, early_stop, reduce_lr],
        verbose         = 1
    )
    
    # ----- ② save EarlyStopping‑restored “final” weights -----------
    final_path = f"ckpts_f3/{tag}_final.weights.h5"
    model.save_weights(final_path)
    print(f"Saved final weights → {final_path}")

  
    model.load_weights(f"ckpts_f3/{tag}_best.weights.h5")
    spice_best = compute_spice(model, val_ds, tok, max_len=max_len, k=1)

    model.load_weights(final_path)  # ES-restored
    spice_final = compute_spice(model, val_ds, tok, max_len=max_len, k=1)

    f3results.append(dict(cfg=cfg,
                        spice_best = spice_best,
                        spice_final= spice_final,
                        val_loss   = min(hist.history['val_loss']),
                        history = hist.history
                        ))
```

```{python, viewimagecaps}
# --- rebuild exactly the same architecture ---
fbest_cfg = winner["cfg"] # already printed above
fbest_cfg = dict(embed_dim=128, lstm_units=512, lr=1e-3)
modelf = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=False)

# choose which checkpoint you want to showcase
ckpt_path = f"ckpts_f3/full1_ed128_lu512_best.weights.h5"
path2 = f"ckpts_f4/best_captioner.h5"
ckpt_path2 = f"ckpts_f2/full1_ed128_lu512_final.weights.h5"
# if you’d rather show the Early-Stopping “final” weights, swap the filename here
modelf.load_weights(ckpt_path2)
print("✓  Weights loaded:", ckpt_path2)

modelf.load_weights(path2)
print("✓  Weights loaded:", path2)


# quick smoke test – run once after training
refstest = {
    0: ["two young boys sitting in a car",      # reference caption 1
        "two children inside a vehicle"]        # reference caption 2
}
refstest = {
    0: ["two children inside a vehicle"]       
}
hypstest = {
    0: ["two children inside a vehicle"]                  # model-generated caption
}

# 2. Instantiate the scorer
scorer = Spice()                 # you can add cache_root=... if you like

# 3. Compute SPICE
corpus_spice, per_image_scores = scorer.compute_score(refstest, hypstest)

print("Corpus-level SPICE:", corpus_spice)
# per_image_scores is a list with one entry per image id



def preprocess_image(path, target_size=224):
    """Read & preprocess 1 image exactly as in training."""
    img_raw = tf.io.read_file(path)
    img     = tf.image.decode_jpeg(img_raw, channels=3)
    img     = tf.image.resize(img, (target_size, target_size)) / 255.0
    return img

import tensorflow as tf
from typing import List, Tuple

def beam_search_decode(model, img_tensor, tokenizer, max_len, k=3, alpha=0.7):
    beam = [([tokenizer.start_id], 0.0)]           # (tokens, log-prob)
    img_batch = img_tensor[None, ...]              # (1, H, W, 3)

    for _ in range(max_len - 1):
        new_beam = []

        for seq, log_p in beam:
            if seq[-1] == tokenizer.end_id:        # branch already finished
                new_beam.append((seq, log_p))
                continue

            # ---------- pad to fixed length ----------
            pad_len  = (max_len - 1) - len(seq)
            seq_pad  = seq + [tokenizer.pad_id] * pad_len
            seq_in   = tf.constant([seq_pad], dtype=tf.int32)   # (1, L)
            preds    = model([img_batch, seq_in])               # (1, L, V)

            # take logits at position len(seq)-1 (current last real token)
            idx_last = len(seq) - 1
            logits   = preds[0, idx_last]
            log_probs = tf.math.log_softmax(logits)

            # ---------- expand top-k ----------
            topk = tf.math.top_k(log_probs, k=k)
            for idx, lp_step in zip(topk.indices.numpy(), topk.values.numpy()):
                new_seq   = seq + [int(idx)]
                new_log_p = log_p + float(lp_step)
                new_beam.append((new_seq, new_log_p))

        new_beam.sort(key=lambda t: t[1], reverse=True)
        beam = new_beam[:k]

        if all(s[-1] == tokenizer.end_id for s, _ in beam):
            break

    # ---------- length-normalised best ----------
    best_seq = max(beam, key=lambda t: t[1] / (len(t[0]) ** alpha))[0]

    # strip special tokens
    if best_seq[0] == tokenizer.start_id: best_seq = best_seq[1:]
    if best_seq and best_seq[-1] == tokenizer.end_id: best_seq = best_seq[:-1]

    return tokenizer.decode(best_seq)



def show_caption(path, caption):
    img_disp = Image.open(path)
    plt.imshow(img_disp); plt.axis('off')
    plt.title(caption, fontsize=12, wrap=True)
    plt.show()
    
    
# ——— tokenizer patch (only needed once) ———
tok.start_id = tok.word2idx["<start>"]
tok.end_id   = tok.word2idx["<end>"]
tok.pad_id   = 0

# grab a handful of validation-set file paths
val_files = val_subset["image"].sample(10, random_state=42)
# (or use test images you’ve never seen)

for fname in val_files:
    path = os.path.join(r.img_dir, fname)
    img_tensor = preprocess_image(path)
    caption    = beam_search_decode(modelf, img_tensor, tok, max_len, k=3)
    show_caption(path, caption)


```

```{python, SPICEmodeltest}

val_spiceset = val_ds.take(10)
modelf = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=False)
#p1 = f"ckpts_f/full1_ed128_lu512_best.weights.h5"
p2 = f"ckpts_f2/full1_ed128_lu512_best.weights.h5"
modelf.load_weights(p2)
print("✓  Weights loaded:", p2)
cp2 = compute_spice(modelf, val_spiceset, tok, max_len=max_len, k=1)
#0.08703

modelf = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=False)
p3 = f"ckpts_f3/full1_ed128_lu512_best.weights.h5"
modelf.load_weights(p3)
print("✓  Weights loaded:", p3)
cp3 = compute_spice(modelf, val_spiceset, tok, max_len=max_len, k=1)
#0.09643

modelf = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=False)
p4 = f"ckpts_f4/best_captioner.h5"
modelf.load_weights(p4)
print("✓  Weights loaded:", p4)
cp4 = compute_spice(modelf, val_spiceset, tok, max_len=max_len, k=1)
#0.09786


```

```{python, retrain}
Path("ckpts_f6").mkdir(parents=True, exist_ok=True)

resultsft = []

modelft = build_model(vocab_size, max_len,
                    embed_dim = 128,        # <- same as best run
                    lstm_units = 512,
                    lr = 1e-3,              # overwritten after loading
                    train_cnn = False)

#ckpt_path = f"ckpts_f3/full1_ed128_lu512_best.weights.h5"
pathretune = f"ckpts_f4/best_captioner.h5"
modelft.load_weights(pathretune)
print("✓  Loaded weights from", pathretune)

#   S = 49 for a 7×7 grid  (or 36 for BUTD regions)
attn_tensor = modelft.get_layer("attn").output[1]   # index 1 = weights
vis_model   = keras.Model(
      inputs  = modelft.inputs,
      outputs = attn_tensor                        # shape (B, T, S)
)
print("✓  Visualiser model ready")

finetune_lr = 1e-5          # start tiny; RLROP will lower further if needed
modelft.compile(optimizer = keras.optimizers.Adam(finetune_lr),
              loss      = "sparse_categorical_crossentropy")

UNFREEZE_AT = -20           # last 20 layers
def unfreeze_cnn(epoch, logs):
    if epoch == 3:
        for layer in modelft.get_layer("mobilenetv2_1.00_224").layers[UNFREEZE_AT:]:
            layer.trainable = True
        print(f"\n🔓  Unfroze last {abs(UNFREEZE_AT)} CNN layers;"
              f" continuing with LR = {keras.backend.get_value(modelft.optimizer.lr):.2e}")

unfreeze_cb = keras.callbacks.LambdaCallback(on_epoch_begin=unfreeze_cnn)

ckpt_fine = keras.callbacks.ModelCheckpoint(
    f"ckpts_f6/finetune_best_spice.h5",
    monitor = "val_spice",
    mode    = "max",
    save_best_only = True,
    save_weights_only = True,
    verbose = 1)

rlrop = keras.callbacks.ReduceLROnPlateau(
    monitor  = "val_loss",
    factor   = 0.5,
    patience = 3,
    min_lr   = 1e-7,
    verbose  = 1)

val_subs = val_ds.take(15)
spice_cb = SpiceCallback(ds       = val_subs, 
                         tok      = tok,
                         max_len  = max_len,
                         every    = 1,
                         k        = 1,
                         out_dir  = "ckpts_f6")

EPOCHS = 10
histft = modelft.fit(
    train_ds,
    epochs          = EPOCHS,
    validation_data = val_ds,
    callbacks       = [ckpt_fine, rlrop, spice_cb, unfreeze_cb],
    verbose         = 1
)

final_path = f"ckpts_f6/retrain_final.weights.h5"
modelft.save_weights(final_path)
print(f"Saved final weights → {final_path}")

resultsft.append(dict(
                    val_loss   = min(histft.history['val_loss']),
                    history = histft.history
                    ))
                    
metrics_history = {
    "loss": [float(x) for x in histft.history["loss"]],
    "val_loss": [float(x) for x in histft.history["val_loss"]],
    "val_spice": [float(x) for x in histft.history["val_spice"]]
}

with open(os.path.join(r.output_dir, "training_metrics.json"), "w") as f:
    json.dump(metrics_history, f, indent=2)

```

```{python, diag plots}
#Ex of first run loss vs val loss
with open("grid_results.json") as h:
    gridrun = json.load(h)

run = gridrun[0]  
exhist = run["hist"]
plt.clf()
plt.plot(exhist["loss"], label="Training Loss")
plt.plot(exhist["val_loss"], label="Validation Loss")
plt.xlabel("Epoch")
plt.ylabel("Loss")
plt.title("Loss Curves for Preliminary Training")
plt.legend()
plt.show()

#After first re-train
# Load the metrics
with open("training_metrics.json") as f:
    metrics = json.load(f)

# Plot loss
plt.figure()
plt.plot(metrics["loss"], label="Training Loss")
plt.plot(metrics["val_loss"], label="Validation Loss")
plt.xlabel("Epoch")
plt.ylabel("Loss")
plt.title("Loss Over Epochs - Re-training")
plt.legend()
plt.show()

# Plot SPICE
plt.figure()
plt.plot(metrics["val_spice"], label="Validation SPICE", color="green")
plt.xlabel("Epoch")
plt.ylabel("SPICE Score")
plt.title("Validation SPICE Over Epochs - Re-training")
plt.legend()
plt.show()

#After second re-train
# Load the metrics
with open("training_metrics2.json") as f:
    metrics2 = json.load(f)

# Plot loss
plt.figure()
plt.plot(metrics2["loss"], label="Training Loss")
plt.plot(metrics2["val_loss"], label="Validation Loss")
plt.xlabel("Epoch")
plt.ylabel("Loss")
plt.title("Loss Over Epochs - Re-training V2")
plt.legend()
plt.show()

# Plot SPICE
plt.figure()
plt.plot(metrics2["val_spice"], label="Validation SPICE", color="green")
plt.xlabel("Epoch")
plt.ylabel("SPICE Score")
plt.title("Validation SPICE Over Epochs - Re-training V2")
plt.legend()
plt.show()



```

```{python, viewcap}
fbest_cfg = dict(embed_dim=128, lstm_units=512, lr=1e-3)
modelre = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=True)

retrain_path = f"ckpts_f6/best_captioner.h5"

modelre.load_weights(retrain_path)
print("✓  Weights loaded:", retrain_path)

# for fname in val_files:
#     path = os.path.join(r.img_dir, fname)
#     img_tensor = preprocess_image(path)
#     caption    = beam_search_decode(modelre, img_tensor, tok, max_len, k=3)
#     show_caption(path, caption)

def show_caption2(path, pred_caption, true_captions=None, max_gt=2):
    """Display an image with its predicted and reference captions."""
    #from IPython.display import display
    img_disp = Image.open(path)
    plt.imshow(img_disp)
    plt.axis('off')

    title = f"Predicted: {pred_caption}"
    if true_captions:
        shown = true_captions[:max_gt]  # only show a few to keep it readable
        gt_text = "\n".join([f"GT {i+1}: {cap}" for i, cap in enumerate(shown)])
        title += f"\n{gt_text}"

    plt.title(title, fontsize=10, wrap=True)
    plt.show()

val_sample = val_subset.sample(10, random_state=42)

for _, row in val_sample.iterrows():
    fname = row["image"]
    gt_captions = row["caption"]  # assumed to be a list of strings
    if isinstance(gt_captions, str):
        gt_captions = gt_captions.split("|")  # adjust delimiter as needed
    path = os.path.join(r.img_dir, fname)

    img_tensor = preprocess_image(path)
    pred_caption = beam_search_decode(modelre, img_tensor, tok, max_len, k=3)

    show_caption2(path, pred_caption, gt_captions)
  
    


```

```{python, re-trainv2}
Path("ckpts_f7").mkdir(parents=True, exist_ok=True)

v2retrain = []

modelft2 = build_model(vocab_size, max_len,
                    embed_dim = 128,      
                    lstm_units = 512,
                    lr = 1e-3,             
                    train_cnn = True)



pathretune2 = f"ckpts_f6/best_captioner.h5"
modelft2.load_weights(pathretune2)
print("✓  Loaded weights from", pathretune2)

cnn = modelft2.get_layer("mobilenetv2_1.00_224")
for layer in cnn.layers:
    layer.trainable = False
print("✓  Re-froze CNN layers for warm-up")

#   S = 49 for a 7×7 grid  (or 36 for BUTD regions)
attn_tensor2 = modelft2.get_layer("attn").output[1]   # index 1 = weights
vis_model2   = keras.Model(
      inputs  = modelft2.inputs,
      outputs = attn_tensor2                        # shape (B, T, S)
)
print("✓  Visualiser model ready")

# ---------------------------------------------------------------
# 2.  Re-compile with a much smaller LR
# ---------------------------------------------------------------
finetune_lr = 1e-5         # start tiny; RLROP will lower further if needed
modelft2.compile(optimizer = keras.optimizers.Adam(finetune_lr),
              loss      = "sparse_categorical_crossentropy")


# UNFREEZE_AT = -10  # last 10 layers
# def unfreeze_cnn(epoch, logs):
#     if epoch == 2:
#         for layer in cnn.layers[UNFREEZE_AT:]:
#             layer.trainable = True
#         print(f"\n🔓  Unfroze last {abs(UNFREEZE_AT)} CNN layers; recompiling...")
# 
#         # Recompile with small LR after unfreezing
#         modelft2.compile(optimizer=keras.optimizers.Adam(1e-5),
#                          loss="sparse_categorical_crossentropy")
# 
# unfreeze_cb = keras.callbacks.LambdaCallback(on_epoch_begin=unfreeze_cnn)

# ---------------------------------------------------------------
# 4.  Callbacks: ModelCheckpoint + ReduceLROnPlateau + SPICE
# ---------------------------------------------------------------
ckpt_fine = keras.callbacks.ModelCheckpoint(
    f"ckpts_f7/finetune_best_spice.h5",
    monitor = "val_spice",
    mode    = "max",
    save_best_only = True,
    save_weights_only = True,
    verbose = 1)

rlrop = keras.callbacks.ReduceLROnPlateau(
    monitor  = "val_loss",
    factor   = 0.5,
    patience = 2,
    min_lr   = 1e-7,
    verbose  = 1)

val_subs = val_ds.take(15)
spice_cb = SpiceCallback(ds       = val_subs, 
                         tok      = tok,
                         max_len  = max_len,
                         every    = 1,
                         k        = 1,
                         out_dir  = "ckpts_f7")

modelft2.fit(train_ds,
             epochs=2,
             validation_data=val_ds,
             callbacks=[ckpt_fine, rlrop, spice_cb],
             verbose=1)


for layer in cnn.layers[-10:]:
    layer.trainable = True
print("🔓  unfroze last 10 CNN layers")


modelft2.compile(optimizer=keras.optimizers.Adam(5e-5),
                 loss="sparse_categorical_crossentropy")


histft2 = modelft2.fit(train_ds,
             initial_epoch=2,   # resumes epoch count
             epochs=10,
             validation_data=val_ds,
             callbacks=[ckpt_fine, rlrop, spice_cb],
             verbose=1)





# EPOCHS = 10
# histft2 = modelft2.fit(
#     train_ds,
#     epochs          = EPOCHS,
#     validation_data = val_ds,
#     callbacks       = [ckpt_fine, rlrop, spice_cb, unfreeze_cb],
#     verbose         = 1
# )

final_path = f"ckpts_f7/retrain_final.weights.h5"
modelft2.save_weights(final_path)
print(f"Saved final weights → {final_path}")


v2retrain.append(dict(
                    val_loss   = min(histft2.history['val_loss']),
                    history = histft2.history
                    ))
                    

metrics_history2 = {
    "loss": [float(x) for x in histft2.history["loss"]],
    "val_loss": [float(x) for x in histft2.history["val_loss"]],
    "val_spice": [float(x) for x in histft2.history["val_spice"]]
}


with open(os.path.join(r.output_dir, "training_metrics2.json"), "w") as f:
    json.dump(metrics_history2, f, indent=2)
    
    
modelft2.save("ckpts_f7/fullmodel_retrain.h5", save_format="h5")
```

```{python, vewcapsv2}
fbest_cfg = dict(embed_dim=128, lstm_units=512, lr=1e-3)
modelrev2 = build_model(vocab_size, max_len, **fbest_cfg, train_cnn=True)

retrain_pathv2 = f"ckpts_f7/best_captioner.h5"

modelrev2.load_weights(retrain_pathv2)
print("✓  Weights loaded:", retrain_pathv2)



import h5py, tensorflow as tf, pprint

vc_path = "ckpts_f7/best_captioner.h5"
modelvc     = modelrev2           # <-- your rebuilt model object

# ------------------------------------------------------------------
# 1️⃣  build a dict of { weight_name : (shape,) } for the model
# ------------------------------------------------------------------
model_shapes = {w.name.split(':')[0]: tuple(w.shape) for w in modelvc.weights}

# ------------------------------------------------------------------
# 2️⃣  build the same dict from the .h5 checkpoint
# ------------------------------------------------------------------
file_shapes = {}
with h5py.File(vc_path, "r") as f:
    def collect(name, obj):
        if isinstance(obj, h5py.Dataset):
            file_shapes[name] = obj.shape
    f.visititems(collect)

# ------------------------------------------------------------------
# 3️⃣  find keys that exist in both but have different shapes
# ------------------------------------------------------------------
mismatch = [(k, model_shapes[k], file_shapes[k])
            for k in model_shapes.keys() & file_shapes.keys()
            if model_shapes[k] != file_shapes[k]]

print("🔎  First five mismatches (if any):")
pprint.pprint(mismatch[:5])
print(f"\nTotal mismatches: {len(mismatch)}")



def show_caption2(path, pred_caption, true_captions=None, max_gt=2):
    """Display an image with its predicted and reference captions."""
    #from IPython.display import display
    img_disp = Image.open(path)
    plt.imshow(img_disp)
    plt.axis('off')

    title = f"Predicted: {pred_caption}"
    if true_captions:
        shown = true_captions[:max_gt]  # only show a few to keep it readable
        gt_text = "\n".join([f"GT {i+1}: {cap}" for i, cap in enumerate(shown)])
        title += f"\n{gt_text}"

    plt.title(title, fontsize=10, wrap=True)
    plt.show()

val_sample = val_subset.sample(10, random_state=42)

val_sample = val_subset.sample(20, random_state=12)
for _, row in val_sample.iterrows():
    fname = row["image"]
    gt_captions = row["caption"]  # assumed to be a list of strings
    if isinstance(gt_captions, str):
        gt_captions = gt_captions.split("|")  # adjust delimiter as needed
    path = os.path.join(r.img_dir, fname)

    img_tensor = preprocess_image(path)
    pred_caption = beam_search_decode(modelft2, img_tensor, tok, max_len, k=3)

    show_caption2(path, pred_caption, gt_captions)

```

```{r, diagnostics}
library(ggplot2)
library(jsonlite)
library(tibble)
res <- fromJSON("grid_results.json", simplifyVector=FALSE)
best <- res[[which.max(sapply(res,"[[","final_spice"))]]
hist <- best$hist
df <- tibble(epoch = as.numeric(seq_along(hist$loss)),
             loss  = as.numeric(hist$loss),
             val_loss = as.numeric(hist$val_loss))
             #spice = as.numeric(hist$val_spice))
# Loss curve
ggplot(df,aes(epoch,loss))+
  geom_line()+geom_line(aes(y=val_loss),linetype=2)+
  labs(title="Training vs validation loss")
# SPICE curve
# ggplot(df,aes(epoch,spice))+geom_line()+
#   labs(title="Validation SPICE across epochs")


```

```{python}
# 1) After training, plot loss curves:
epochs = range(1, len(hist.history['loss']) + 1)

plt.figure()
plt.plot(epochs, hist.history['loss'], label='train loss')
plt.plot(epochs, hist.history['val_loss'], label='val loss')
plt.xlabel('Epoch')
plt.ylabel('Loss')
plt.title('Training & Validation Loss')
plt.legend()
plt.show()

# 2) After your debug forward‐pass, plot the token‑ID histogram:
#    (you’ve already got token_ids from `np.argmax(logits,axis=-1)`)
flat_ids = token_ids.flatten()
unique, counts = np.unique(flat_ids, return_counts=True)

plt.figure()
plt.bar(unique, counts)
plt.xlabel('Token ID')
plt.ylabel('Count')
plt.title('Predicted Token Distribution')
plt.show()



```

